An efficient acyclic contact planner for multiped robots

### 2016-01-20 body/abstract.tex
typos

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in an arbitrary cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoretical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the feasibility
of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts. For each limb, an off-line sampling approach,
independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the equilibrium of the robot.
These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
variety of robots, including HRP-2, HyQ, and a dexterous hand, with interactive performances.

>>>>

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in a cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoretical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the feasibility
of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts.
A sampling based-approach is used to generate candidate configurations.
The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the equilibrium of the robot.
These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
variety of robots, including HRP-2, HyQ, and a dexterous hand, with interactive performances.

### 2016-01-20 body/efort.tex
typos

Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
the equivalent joint torque vector $\mathbf{\tau}$;
we can define the mechanical work in both spaces:
\begin{equation*} \label{power}
\dot{\mathbf{q}}^T \mathbf{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
\end{equation*}

Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
\begin{equation*} \label{ballforce}
||\mathbf{\tau}||^2 \leq 1
\end{equation*}
is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
\begin{equation} \label{ellipsoidforce}
\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
\end{equation}

@@

With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
We thus consider the force ellipsoid as a basis for our heuristic.
In a given direction $\mathbf{m}$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
\begin{equation*}
f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(JJ^{T})\mathbf{m}]^{-\frac{1}{2}}
\end{equation*}

>>>>

Similarly to the velocity ellipsoid, \citeauthor{Yoshikawa1984} also defines the force ellipsoid.
Considering: a force vector $\mathbf{f}$ expressed in the task space $\mathbb{R}^m$;
the equivalent joint torque vector $\bm{\tau}$;
we can define the mechanical work in both spaces:
\begin{equation*} \label{power}
\dot{\mathbf{q}}^T \bm{\tau} = \dot{\mathbf{p}}^T \mathbf{f}
\end{equation*}

Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
\begin{equation*} \label{ballforce}
||\bm{\tau}||^2 \leq 1
\end{equation*}
is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
\begin{equation} \label{ellipsoidforce}
\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
\end{equation}

@@

With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
We thus consider the force ellipsoid as a basis for our heuristic.
In a given direction $\mathbf{m}$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
\begin{equation*}
f_\mathsf{T}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}	^{T})\mathbf{m}]^{-\frac{1}{2}}
\end{equation*}

### 2016-01-20 body/glossary.tex
typos

\newglossaryentry{reachc}
{
  name=\textit{reachability condition},
  description={A root configuration $\mathbf{q}^{0}$ that verifies :\\ $W(\mathbf{q}^{0}) \cap O \neq \emptyset \wedge W^0_{s^*}(\mathbf{q}^{0}) \cap O = \emptyset$,
  where $s^*$ is a user-defined value}
}

>>>>

\newglossaryentry{reachc}
{
  name=\textit{reachability condition},
  description={A root configuration $\mathbf{q}^{0}$ that verifies :\\ $W(\mathbf{q}^{0}) \cap O \neq \emptyset \text{ and } W^0_{s^*}(\mathbf{q}^{0}) \cap O = \emptyset$,
  where $s^*$ is a user-defined value}
}

### 2016-01-20 body/intro.tex
typos

As suggested in \cite{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner, while preserving the completeness of the search.
We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot path.
This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly.
while preserving a high success rate in the demonstrated scenarios.

@@

For completeness, we lastly mention a new kind of approach, recently proposed from the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
use black box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
at \gls{interactive} frame rates. While this new approach provides an exciting direction of research, currently the resulting motions
look unnatural, and do not seem applicable to real robots.

Regarding robotics applications, it thus appears that optimization based approaches outperform existing planners,
partly because the interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem.
However recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on the real robot.
To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. Such a planner holds the promise of a complete real-time multi-contact locomotion system.

@@

In Section~\ref{overview}, we present the general organization of our method. Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb. In Section~\ref{sec:heuristics}, we present heuristics used for the selection of a contact configuration. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
that complete those presented in the conference paper.
\subsection*{Comparison with our previous work.}
The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work and brings new contributions, resulting from the implementation of our approach real-world robots and problems.
In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
As opposed to the conference paper, the complete algorithm for the contact planner is provided, and an access to the source code of our implementation is provided.
Furthermore, our experiments are supported by a an in-depth analysis of our performance, as well as a discussion on the influence of the parameters to the method.

>>>>

As suggested in \cite{Bouyarmane2009}, we believe that these two tasks can be treated sequentially within a global planner.
We go further and claim that this can be done at a much smaller cost, provided that necessary and sufficient conditions for contact feasibility can be formulated for the robot path.
This paper presents a geometrical representation of these conditions, and a concrete implementation of such a decoupled planner.
This implementation results from a trade-off between a necessary and a sufficient condition for feasibility, allowing us to find solutions extremely rapidly.
while preserving a high success rate in the demonstrated scenarios.

@@

For completeness, we lastly mention a new kind of approach, recently proposed from the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
use black box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
at \gls{interactive} frame rates (We say that a planning method is interactive when the computation time for one step is lesser than the
time to execute it). While this new approach provides an exciting direction of research, currently the resulting motions
look unnatural, and do not seem applicable to real robots.

Regarding robotics applications, it thus appears that optimization based approaches outperform existing planners,
partly because the real-time interpolation between contact poses ($\mathcal{P}_3$) remains an unsolved problem.
However recent contributions have been able to bring partial solutions to it \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}. \citeauthor{Carpentier2016} are able to achieve this with \gls{interactive} performances on the real robot.
To generate the input contact configurations, there is a need for an efficient contact planner, able to break the combinatorial to generate discrete contact sequences rapidly. Such a planner holds the promise of a complete real-time multi-contact locomotion system.

@@

In Section~\ref{overview}, we present the general organization of our method. Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb. In Section~\ref{sec:heuristics}, we present heuristics used for the selection of a contact configuration. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
that complete those presented in the conference paper.
\subsection*{Comparison with our previous work.}
The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work and brings new contributions, resulting from the implementation of our approach real-world robots and problems.
In particular, a strong contribution is the introduction of a robust equilibrium criterion, designed to ensure the equilibrium of the robot despite bounded errors in the contact forces, coupled
with heuristics for relevant contact selection. These two contributions were required to apply our method to real-robot models, namely HRP-2 and HyQ~\citep{semini11hyqdesignjsce}.
In the present paper, a complete algorithm for the contact planner is given, and an access to the source code of our implementation is provided.
Furthermore, our experiments are supported by a an in-depth analysis of our performance, as well as a discussion on the influence of the parameters to the method.

### 2016-01-20 body/overview.tex
typos

\begin{figure}[t]
\centering
  \begin{overpic}[width=1\linewidth]{figures/2D_feas}
		\put (15,) {$C_{Equil}^0$      $\subset$} 
		\put (47,) {$C_{Contact}^0$ $\approx$ } 
		\put (76,) {$C_{Reach}^0$} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Illustration of several configurations sets used in this paper in a 2D scene. Obstacles are violet, and units are in meters. To show the sets in a 2D representation, all the rotational joints of HRP-2 are locked in the shown configuration, such that a torso configuration
is only described by two positional parameters (x and y). $C_{Equil}^0$ is included in $C_{Contact}^0$. $C_{Reach}^0$ approximates $C_{Contact}^0$. Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$. Considering the configurations around the top obstacle, we can observe a dramatic
divergence between  $C_{Equil}^0$  and $C_{Contact}^0$ when the problem is not \textit{\gls{cluttered}}.}
		   \label{fig:dedefeas}
\end{figure}

@@

We make the assumption that for the \textit{\gls{cluttered} contact planning problem} (that we define as admitting a solution where every configuration contains at least one contact
is \gls{quasi-flat}) verifying C2 is not critical for $\mathcal{P}_1$.
This means that root configurations satisfying C1 and C3 also satisfy C2. For this reason we neglect C2 during the resolution of \Pa, and we worry about it only when solving \Pb.

@@

An intuitive description of \gls{contact feasible} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:dedefeas} -- Red).
Our objective is to find a root path such that the root (scaled by a factor $s>1$) is not in collision, while the reachable workspace is in collision with the environment.
In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is designed and sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).

@@

From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
The algorithm is designed so that only one contact can be created or broken between two successive configurations. While does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
Details are presented in Section~\ref{sec:contact}.

@@

\medskip
\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains.
Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

>>>>

\begin{figure}[t]
\centering
  \begin{overpic}[width=1\linewidth]{figures/2D_feas}
		\put (15,) {$C_{Equil}^0$      $\subset$} 
		\put (47,) {$C_{Contact}^0$ $\approx$ } 
		\put (76,) {$C_{Reach}^0$} 
		%~ \put (5,27) {3.b)} 
		%~ \put (37,27) {4.a)} 
		%~ \put (68,27) {4.b)} 
	\end{overpic}
\caption{Illustration of several root configurations sets used in this paper in a 2D scene. Obstacles are violet, and units are in meters. To show the sets in a 2D representation, all the rotational joints of HRP-2 are locked in the shown configuration, such that a torso configuration
is only described by two positional parameters (x and y). $C_{Equil}^0$ is included in $C_{Contact}^0$. $C_{Reach}^0$ approximates $C_{Contact}^0$. Depending on a parametrization, we can obtain $C_{Contact}^0 \subset C_{Reach}^0$. Considering the configurations around the top obstacle, we can observe a dramatic
divergence between  $C_{Equil}^0$  and $C_{Contact}^0$ when the problem is not \textit{\gls{cluttered}}.}
		   \label{fig:dedefeas}
\end{figure}

@@

We make the assumption that for the \textit{\gls{cluttered} contact planning problem} (that we define as admitting a solution where every configuration contains at least one contact
is \gls{quasi-flat}) verifying C2 is not critical for $\mathcal{P}_1$.
This means that root configurations satisfying C1 and C3 also satisfy C2. For this reason we neglect C2 during the resolution of \Pa, and we consider it only when solving \Pb.

@@

An intuitive description of \gls{contact feasible} configurations is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right leg in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision.
We approximate $C_{Contact}^0$ with a set $C_{reach}^0$, defined by the \textit{reachability condition}, a geometrical criterion based on this intuitive description (Figure~\ref{fig:dedefeas} -- Red).
Our objective is to find a root path such that the root (scaled by a factor $s \geq 1$) is not in collision, while the reachable workspace is in collision with the environment.
In Section~\ref{rbprm} we detail how  $C_{reach}^0$ is designed and sampled to compute a guide path with a dedicated planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa).

@@

From a given start configuration, the planner proceeds in an iterative fashion along the discretized path: given a new root configuration, an inverse-kinematics solver 
is used to maintain the contacts that were previously existing. Possibly, some of these contacts cannot be maintained (because of joint limits or collisions).
Then the contacts are broken. Conversely, new contacts are created to ensure the static equilibrium of the robot.
The algorithm is designed so that only one contact can be created or broken between two successive configurations. While it does not provide guarantees that the interpolation between two configurations is achievable, it appears empirically as a reasonable heuristic.
Details are presented in Section~\ref{sec:contact}.

@@

\medskip
\textbf{The environment} $O$ is defined as the union of the obstacles $O_i$ that it contains.
%~ Finally, we define $dist(\mathbf{x}, O)$ as the minimal distance between a point $\mathbf{x}$ and the closest surface of the environment.

### 2016-01-20 body/results.tex
typos

\begin{table}
\centering
\footnotesize
\begin{tabular}{c | c | c}
   Value of $s$ &  False positive & False negative\\
 \hline
   1   & 24\% & 0 \%\\
   1.1& 12\% & 4\% \\
   1.15& 07\% & 6\%\\
   1.2 & 03\% & 7.5\%\\
   1.25& 02\% & 8.3\%\\
   1.5 & 01\% & 9.5\%\\
 \end{tabular}
\caption{Percentage of true and false positive in the reachability condition, depending on the scaling value of $W^0$.}
\label{tab:scale}
\quad
\end{table}

\subsubsection{On the choice of heuristics} \label{sec:heuristics}
In our conference paper, the computed motions were only generated using the EFORT heuristic.
EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
Regarding locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
configurations close to singularities (and joint limits). While this does not significantly impact
the generation of the plan, the resulting interpolation on the real robot turned out to be harder \citep{Carpentier2016}.
For this reason, we now prefer to use our manipulability based heuristic for the legs of the robot, but we still
use EFORT for the arm, that results in lesser contact repositionning.

>>>>

\begin{table}
\centering
\footnotesize
\begin{tabular}{c | c | c}
   Value of $s$ &  False positive & False negative\\
 \hline
   1   & 24\% & 0 \%\\
   1.1& 12\% & 4\% \\
   1.15& 7\% & 6\%\\
   1.2 & 3\% & 7.5\%\\
   1.25& 2\% & 8.3\%\\
   1.5 & 1\% & 9.5\%\\
 \end{tabular}
\caption{Percentage of true and false positive in the reachability condition, depending on the scaling value of $W^0$.}
\label{tab:scale}
\quad
\end{table}

\subsubsection{On the choice of heuristics} \label{sec:heuristichoices}
In our conference paper, the computed motions were only generated using the EFORT heuristic.
EFORT is designed for tasks requiring to exert important forces (such as pushing / pulling / climbing). 
Regarding locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
configurations close to singularities (and joint limits). While this does not significantly impact
the generation of the plan, the resulting interpolation on the real robot turned out to be harder \citep{Carpentier2016}.
For this reason, we now prefer to use our manipulability based heuristic for the legs of the robot, but we still
use EFORT for the arm, that results in lesser contact repositionning.
